The Regional Effects of Marine Protected Areas

### 2019-10-10 documents/ovando-regional-effects-of-mpas.tex
initial submission

\usepackage{eso-pic,graphicx,transparent}

@@

  \affil[a]{University of Washington, School of Aquatic and Fishery Science}
  \affil[b]{University of California, Santa Barbara}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Healthy marine ecosystems are critical to the well-being of the planet.
Marine protected areas, parts of the oceans protected from human
activities such as fishing, are increasingly being used in an effort to
conserve and manage these ecosystems. Our study pairs theory with
empirical methods to examine the regional effects of MPAs on
conservation and fishery outcomes. Using simulations, we show that the
effects of marine protected areas can be highly variable and dependent
on human drivers. Building off of MPA theory, we find no statistically
clear effect of a network of MPAs, which our simulation analysis
suggests is entirely reasonable. Our results lay a foundation for future
research on the design and monitoring of marine protected areas.}

@@

\authordeclaration{R.H's research program receives funding from environmental NGOs,
foundations, fishing industry, governments and international agencies.
All of these can be interpreted as a conflict of interest when
evaluating fisheries policy. C.C., S.G., and O.D's research program
includes funding from environmental NGOs and foundations with an
interest in ocean conservation and food security}

@@

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
 \keywords{  Marine Protected Areas |  Conservation |  Bio-economic modeling |  Program Evaluation |  Channel Islands National Marine Sanctuary  }

@@

\acknow{This study would not be possible without the work provided by PISCO
divers over the years. DO thanks Cody Szuwalski, Julia Lawson, and André
Punt for helpful comments and technical support.}

@@

No-take Marine Protected Areas (MPAs), spatial regions of the ocean in
which fishing is prohibited, have a long history in the management of
marine resources. Traditional cultures in Oceania utilized - often
temporary - MPAs as ``fish banks'' for times of need (1). Modern MPAs
were first established primarily as marine analogs to the terrestrial
protection of iconic landscapes like Yellowstone or Kruger National
Parks (2, 3). Over time our goals and expectations for MPAs have
evolved; while all MPAs are expected to deliver conservation benefits
within their borders, many modern MPAs are also established to bolster
fish populations throughout the region in which they are located (what
we term ``regional-scale effects'') (4).

@@

Much of the literature on the effects of Marine Protected Areas focuses
on assessing the conservation effects within the borders of protected
areas (7). While these within-MPA effects are vitally important for
protecting rare species, biodiversity, critical habitats, and often
tourism, they paint an incomplete picture of the overall population
effects of MPAs. The organisms within the borders of protected areas are
generally part of a broader population, the biological stock, connected
through adult movement and larval dispersal. If the goal of
conservationists or natural resource managers is to increase the total
abundance or productivity of a resource, a broader question we should
ask of MPAs is not just are there more fish inside their borders, but
also how have the reserves affected abundances throughout the region in
which they are located? This logic, that MPAs will have conservation
benefits for most species beyond their borders, is implicit in all
multilateral calls for MPA expansion.

We define the regional conservation effects of MPAs as the change in
total biomass of fish (summing inside and outside of MPAs) relative to
the total biomass of fish that would have occurred without the MPAs
(acknowledging that other outcomes such as increased biodiversity or
resiliency are also important to conservation but are beyond the scope
of this analysis). Note that this definition of regional effects is in
line with the recommendations of (22), but differs from effects measured
by for example differences in outcomes before and after MPAs.

We used a simulation model to examine the regional conservation effects
of MPAs under a wide array of conditions that might reasonably be
expected to affect these outcomes. The model explores combinations of
MPA designs (e.g., sizes, number, and placement of MPAs), life histories
(e.g., growth and mortality rates, adult movement, larval dispersal,
recruitment variability and autocorrelation, and timing of density
dependence), and fishing fleet dynamics (e.g., degree of fishing
pressure, size selectivity of the fleet, reaction of fishing pressure to
MPA placement) (See Table.S1 for a complete description of simulation
variables). We use this model to explore how different drivers interact
to affect the conservation and fishery outcomes of MPAs.

@@

As a thought experiment, imagine a region that has driven its fish
populations to near extinction and then places 100\% of its waters
inside a no-take MPA. In that setting, we would expect the
regional-scale conservation benefits to be massive. On the other hand,
imagine implementing an MPA in an area where a sedentary species has
been only lightly fished. In that setting, it is possible to create a
small net conservation loss if the concentration of fishing pressure
outside the reserve has a greater effect than the biomass buildup inside
the reserve. Within these broad bounds, numerous factors can act to
affect the regional effects of MPAs. These include the scale of adult
and larval dispersal relative to the size of the MPAs ({\textbf{???}},
23--25), the strength and timing of density dependence in the population
(e.g.~pre- or post-settlement), how overfished the population would be
without the MPA, and how fishing activity and management responds to the
implementation of the MPAs (4, 8, 26--32). In addition, even for the
same total area of MPAs, the location and spacing of the MPAs can have a
profound influence on their cumulative impact through habitat and
network effects (4, 33).

@@

These simple results support the intuitive conclusion that bigger MPAs
on overfished populations produce large positive conservation outcomes.
However, when combined with other factors such as fleet dynamics, life
history, and MPA size, the conservation effects of MPAs can vary widely
in magnitude and timing, a finding corroborated by empirical evidence
(37--40) (Fig.\ref{conservation-effects}-B). MPAs protecting small
proportions of a stock can produce small positive effects, large
positive effects, or even small negative effects. As the area protected
inside MPAs increases, positive effects become more likely, but even
MPAs protecting most of a population's footprint can produce
conservation effects as low as 0\% and as high as 200\% and above
depending on the context in which they are placed. The degree of
depletion (i.e.~fishing pressure) has a clearer signal, with small
conservation benefits of MPAs when depletion is low, and larger effects
when depletion is high, but again even for severely depleted populations
the conservation effects of MPAs can vary widely (although nearly always
with population benefits).

@@

How can we empirically measure the regional effects of MPAs? Intuition
and bio-economic theory tells us that the regional conservation effects
of MPAs are likely to be highly context dependent across locations and
across species within locations, so careful thought must be devoted to
an empirical design to detect their effects. The perfect experiment
would involve two parallel worlds that were identical, except for the
implementation of an MPA. In world ``A'', no MPA would be implemented,
and in the facsimile world ``B'', the MPA would be implemented. Both
worlds would be tracked before and after MPA implementation, and the
conservation and fishery outcomes would be compared after treatment.
Instead of two parallel worlds, a similar experimental design would
involve random placement of MPAs. Unfortunately, neither of these
experimental designs has been implemented, because parallel worlds do
not exist and MPAs are never placed at random. Careful causal inference
requires some other means of controlling for biases introduced by
factors such as the MPA siting process and biological spillover and
concentration of fishing effort (22, 41).

Conservation effects are typically estimated by comparing organism
densities or biomass inside MPAs to densities or biomass in selected
control sites outside MPAs, which we will refer to as response ratios.
(5) and (42) present meta-analyses of hundreds of such studies. These
results often find massively higher densities and biomasses inside MPAs
than outside (5). While response ratios are intended as a measure of the
effectiveness of reserves within their borders, these result stand in
contrast to the smaller region wide effect sizes predicted by our
simulation model. This raises the question: How reliable are density or
biomass ratios as estimators of regional conservation effects?

Control sites used in density or biomass ratios are often selected based
on abiotic or ecological traits such as habitat characteristics (22).
However, selection of control sites is further complicated by the very
spillover that MPAs are often intended to create. Export of adults or
larvae from the MPA to the ``control'' site affects their status as
controls, as does displacement of fishing effort from MPAs to control
sites. In theory, control sites far enough away to negate both
biological spillover and concentration by the fishing fleet could be
selected, but finding suitably far sites that are also appropriate
proxies for the ecological and economic context of the MPAs may be
challenging. While these concerns have been stated previously (43), the
MPA evaluation literature has by and large been unable to adequately
address them, often due to the very real challenges of acquiring the
necessary data (22).

Failure to account for these spillover effects can result in a biased
estimate of the true effect of a policy such as MPA placement (41). In
an ideal setting, the control site used in a response ratio would be a
perfect counterfactual for what would have occurred without the MPA. We
used our simulation model to approximate this scenario, calculating the
response ratio as the ratio of densities inside the MPAs relative to the
overall density from the paired simulation without MPAs multiplied by
the area of the population in an MPA (to account for the fact that we
would expect the effectiveness of a response ratio as a proxy for
regional conservation effects to scale with the size of the MPA relative
to the range of the species, Fig.\ref{density-ratio}-A). In this
idealized case, the response ratio is often a reasonable estimator for
the regional conservation effect.

What happens when the ``control'' sites are in fact affected by the
application of the MPA through biological spillover and fishing
concentration? In order to address this possibility, we calculated a new
response ratio as the mean density inside MPAs relative to the mean
density outside the MPAs (both weighted by distance from MPA borders and
scaled by MPA size). We only include simulations in which habitat and
larval dispersal rates are identical inside and outside of MPAs, to
approximate a scenario in which treatment and control sites have been
paired by ecological characteristics. Under these circumstances the
response ratio is a somewhat biased and very inaccurate estimator of the
true regional effect of MPAs. While on average high response ratios
correspond with high true regional effects, the variance is high, so the
response ratio alone cannot be used as a reliable proxy for the true
regional effect, without accounting for the sources of biases discussed
here. For higher movement rates, response ratios were frequently near
0\% (potentially leading stakeholders to conclude that the MPA had been
ineffective), when in fact in many cases the true effect of the MPA was
highly positive. For highly sedentary species, extremely high response
ratios could create the appearance of massive conservation gains when in
fact the net effect on the regional population has been near zero or
even negative (Fig.\ref{density-ratio}-B).

@@

To the extent that we can select control sites that are both
sufficiently similar to the region treated with MPAs and are unaffected
by biological or economic spillover, response ratios may be reliable
estimators of regional conservation effects. In reality these conditions
are unlikely to hold but for a few select cases. This become especially
difficult when networks of MPAs are implemented, such as has occurred in
California. In this case, control sites are generally located within a
few kilometers from MPA borders, making biological and economic
spillover highly likely. In these circumstances, without robust
statistical controls, response ratios may provide an unreliable estimate
of the regional conservation effect of marine protected areas.

\begin{figure*}%[tbhp]
  \centering
  \includegraphics[width=.9\linewidth]{figs/density-ratio.pdf}
  \caption{Simulated MPA-size weighted response ratios (x-axis) plotted against true regional change in biomass caused by MPAs (y-axis). Each point represents a simulation, and color represents the movement rate of fish in the simulation (expressed as a percentage of the patches a fish is likely to move in one time step). The black line shows the one:one fit line, the blue line is a linear fit of the relationship between observed response ratios and true regional change in biomass. Inside vs. Experiment simulates a true control, while Inside Vs. Outside uses a control that may be affected by the treatment.}
  \label{density-ratio}
\end{figure*}

@@

Given these challenges, what is a viable alternative for estimating the
conservation effects of MPAs? In a manner similar to (38), we use a
difference-in-difference estimator comparing densities in finfish
species targeted by fishing effort (i.e., those potentially affected by
an MPA) from those not targeted before and after MPA implementation
(41). Targeted species in the Northern Channel Islands, California,
include commercially important finfish such as California sheephead
(\emph{Semicossyphus pulcher}), and copper (\emph{Sebastes caurinus})
and blue (\emph{Sebastes mystinus}) rockfish. Each of these targeted
species was the subject of prior bio-economic modeling related to the
effects of MPAs in southern California (44, 45). But, we must note that
the analysis omits species from important invertebrate fisheries
including red urchin (\emph{Mesocentrotus franciscanus}) and spiny
lobster (\emph{Panulirus interruptus}). Non-targeted species include
garibaldi (\emph{Hypsypops rubicundus}), halfmoons (\emph{Medialuna
californiensis}), and blacksmith (\emph{Chromis punctipinnis}) (See
Table.S2 for a complete list of species). The key assumptions of this
approach are that a) within the time-frame of the model there are no
significant interaction effects between the targeted and non-targeted
species (which in fact we do not detect, see SI section 1.6.3), and b)
that in the absence of the MPAs both the targeted and non-targeted
groups of species would have exhibited similar trends in densities (the
parallel-trends assumption). The advantage to this approach is that
given the time-frame of the model (15 years), we believe that spillover
effects of MPA placement on the non-targeted species are likely to be
much less severe than the effects of biological spillover and fishing
concentration that may bias the performance of estimators such as a
response ratios or a spatial BACI design.

We applied this difference-in-difference strategy to a large dataset in
California. We used empirical kelp forest survey data from the
Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO)
monitoring in the Northern Channel Islands with the ultimate goal of
testing the regional effects of MPAs in a real world context. A network
of MPAs covering approximately 20\% of the islands' waters was put in
place in 2003 as part of the California Marine Life Protection Act
(MLPA) (see (46), (47), (48), and (49) for information on the creation
of the MLPA). PISCO conducts visual SCUBA surveys at a large number of
rocky-reef and kelp forest sites inside and outside of MPAs throughout
the Channel Islands, producing estimates of densities of fishes that are
both targeted and non-targeted by fishing
(Fig.\ref{channel-islands}-\ref{pop-trends}).

@@

\begin{figure}%[tbhp]
  \centering
  \includegraphics[width=1\linewidth]{figs/pop-trends.pdf}
  \caption{Centered and scaled mean annual biomass densities of included targeted (red) and non-targeted (blue) fish species (faded lines) and smoothed means with 95\% confidence intervals over time (darker lines and ribbons).}
  \label{pop-trends}
\end{figure}

The raw trends in the biomass densities appear to show parallel trends
in the densities of targeted and non-targeted fishes pre-MPA, with the
targeted group diverging first positively and then negatively over time
(Fig.\ref{pop-trends}). However, these raw trends could be caused by any
number of confounding variables besides the effects of the MPAs
themselves, from changes in observer skill and sampling locations, to
shifts in environmental conditions. Using our difference-in-difference
model in an effort to control for these factors, we found no significant
difference in the biomass densities of targeted and non-targeted fishes
pre-MPA (before 2003), which provides support for the parallel trends
assumption critical to the difference-in-difference method. Following
the implementation of the MPAs in 2003, we see evidence of an increasing
trend in densities of targeted relative to non-targeted fish, reaching a
peak value in 2014. Following 2014, densities of targeted species appear
to have declined relative to what we would expect based on the
non-targeted group (Fig.\ref{did-plot}).

\begin{figure}%[tbhp]
  \centering
  \includegraphics[width=1\linewidth]{figs/did_plot.png}
  \caption{Estimated divergence in biomass densities of targeted and non-targeted fishes throughout the Channel Islands (integrated across inside and outside of MPAs). MPAs are implemented in 2003 (red dashed line). Estimates are from a regression on log(biomass density), so estimated effects roughly correspond to percentage changes. Colored cells in the background indicate the frequency of MPA conservation effects predicted by our simulation modeling for MPAs similar in design and species composition to the Channel Islands.}
  \label{did-plot}
\end{figure}

@@

We find regional conservation gains from MPAs in 94\% of our simulations
(Fig.\ref{conservation-effects}). MPAs covering less than 25\% of the
population range were unlikely to produce regional conservation gains
above 10\% unless the stock would have been severely overfished without
the MPA. Given that few marine species on the planet have more than 25\%
of their entire range protected in MPAs, this suggests that large
regional gains are unlikely to be common unless the overall extent of
MPAs grows globally. Large MPAs protecting highly depleted stocks almost
always produced large regional conservation gains, as should be
expected. The median conservation effect of 36\% stands in contrast to
the large within-MPA effects on biomass densities reported by (50) and
(5). This difference is due in part to the fact that within-MPA effects
are likely to be much larger than regional-MPA effects, but may also
reflect the many challenges of translating response ratios into regional
conservation effects. Our simulation analysis shows that without proper
controls (in particular for the effects of fleet redistribution),
response ratios well over 100\% are entirely plausible even when the
actual conservation effect is much lower (and \emph{vice versa}).

@@

However, our simulation analysis suggests that this result should
perhaps be expected. The Channel Islands MPAs cover approximately 20\%
of the waters in the Channel Islands, and while formal stock assessments
are not available for many of the targeted species in our analysis, what
evidence we have suggests that, as a group, these fish are on average
not heavily overfished. Some species, such as California sheephead and
blue rockfish were likely below target levels during the period (51,
52), but projections based upon the overall average response across all
species will likely suggest modest benefits even if a subset of species
could experience much larger population gains. Harkening back to our
simulations, we expect the average percentage difference in densities of
targeted species with and without MPAs to be modest
(Fig.\ref{did-plot}). Effects of this size are likely to be challenging
to detect empirically given the large natural variation of marine
ecosystems (especially temperate reefs) and the observation error
inherent in infrequent annual survey programs such as those provided by
PISCO.

@@

While the MPA effect estimated by our model was not significant in any
one year, the positive trend from 2005 to 2014 is evident, as is the
sharp decline from 2015-2017. To what should we attribute this apparent
shift? We cannot reject the parallel trends assumption between the
targeted and non-targeted species in the years before the MPAs
(Fig.\ref{did-plot}). But, the Channel Islands region (and the entire
West coast of the USA) experienced a dramatic `marine heatwave'
beginning in 2014 and persisting through 2016, resulting in part in
extremely elevated water temperatures throughout the region (53).
Biogeographic differences in the distributions of targeted and
non-targeted species may confound the observed effects of MPAs. Many of
the non-targeted species have warm thermal affinities and have increased
in numbers since the heatwave (54). However, the targeted group is made
up mostly of fishes with cold-water affinities, such as members of the
genus Sebastes. As such, we hypothesize that the recent evidence for a
decline in densities of targeted species is due to environmental
conditions that disproportionately affect the targeted group (and not
for example due to concentrated fishing pressure outside the reserves).
This hypothesis is supported by the sharp declines in densities of
targeted species seen inside the MPAs themselves (see Fig.S27-S29),
suggesting a driver other than fishing may be at play. The Channel
Islands case study demonstrates the challenge of detecting small effect
sizes in nature and highlights the potential for small effect sizes to
be overwhelmed by additional environmental shocks to the system.

@@

Once simulation models have been used to help design an MPA, how can
users evaluate whether it is achieving their objectives? Response ratios
are commonly used as evidence for conservation outcomes of MPAs (5), but
as suggested in (22) and (41) and demonstrated here, without careful
attention to the design of control sites (accounting for example for the
displacement of effort by MPAs), response ratios may be highly
unreliable estimators of regional MPA effects. As (41) suggests, there
are many potential alternatives for estimating the effects of MPAs that
better account for the challenges of causal inference. We applied one
such approach here (a difference-in-difference estimator), and yet were
still unable to reach robust conclusions as to the effect of MPAs on the
density of targeted species in the Channel Islands, due to the likely
small size of the true effect relative to the strength and variability
of environmental drivers.

>>>>

\usepackage{eso-pic,graphicx,transparent,xcolor}

@@

  \affil[a]{University of Washington, School of Aquatic and Fishery Sciences}
  \affil[b]{University of California, Santa Barbara}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Healthy marine ecosystems are critical to the well-being of the planet.
Marine protected areas, parts of the oceans protected from human
activities such as fishing, are increasingly being used in an effort to
conserve and manage these ecosystems. Our study pairs theory with
empirical methods to examine the regional effects of MPAs on
conservation outcomes. Using simulations, we show that the regional
effects of marine protected areas can be highly variable and dependent
on human drivers. We find no statistically clear effect of a network of
MPAs after 13 years of protection, which our simulation analysis
suggests is actually to be expected. Our results lay a foundation for
future research on the design and monitoring of marine protected areas.}

@@

\authordeclaration{R.H's research program receives funding from environmental NGOs,
foundations, fishing industry, governments and international agencies.
All of these can be interpreted as a conflict of interest when
evaluating fisheries policy. C.C., S.G., and O.D's research program
includes funding from environmental NGOs and foundations with an
interest in ocean conservation and food security. Funding for D.O. was
partly provided by the NMFS-Sea Grant Population and Ecosystem Dynamics
Fellowship.}

@@

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
 \keywords{  Marine Protected Areas |  Conservation |  Bio-economic modeling |  Program Evaluation  }

@@

\acknow{Funding for the simulation model was provided by the NMFS-Sea Grant
Population and Ecosystem Dynamics Fellowship. This study would not be
possible without the work provided by PISCO divers over the years. DO
thanks Cody Szuwalski, Julia Lawson, and André Punt for helpful comments
and technical support.}

@@

No-take Marine Protected Areas (MPAs), spatial regions of the ocean in
which fishing is prohibited, have a long history in the management of
marine resources. Traditional cultures in Oceania utilized - often
temporary - MPAs as ``fish banks'' for times of need (1). Modern MPAs
were first established primarily as marine analogs to the terrestrial
protection of iconic landscapes like Yellowstone or Kruger National
Parks (2, 3). Over time our goals and expectations for MPAs have
evolved; while all MPAs are expected to deliver conservation benefits
within their borders, many modern MPAs are also established to bolster
fish populations throughout the region in which they are located (4).

@@

Much of the literature on the effects of MPAs focuses on assessing the
conservation effects within the borders of protected areas (7). While
these within-MPA effects are vitally important for protecting rare
species, biodiversity, critical habitats, and often tourism, they paint
an incomplete picture of the overall population effects of MPAs. The
organisms within the borders of protected areas are generally part of a
broader biological stock, connected through adult movement and larval
dispersal. If the goal of conservationists or natural resource managers
is to increase the total abundance or productivity of a resource, a
broader question we should ask of MPAs is not just are there more fish
inside their borders, but also how have the reserves affected abundances
throughout the region in which they are located? This logic, that MPAs
will have conservation benefits for most species beyond their borders,
is implicit in all multilateral calls for MPA expansion.

We define the regional conservation effects of MPAs as the change in
total biomass of fish (summing inside and outside of MPAs) relative to
the total biomass of fish that would have occurred without the MPAs
(acknowledging that other outcomes such as increased biodiversity or
resiliency are also important to conservation but are beyond the scope
of this analysis).

We used a simulation model to examine the regional conservation effects
of MPAs under a wide array of conditions that might reasonably be
expected to affect these outcomes. The model explores combinations of
MPA designs, life histories, and fishing fleet dynamics (See Table.S1
for a complete description of simulation variables). We use this model
to explore how different drivers interact to affect the conservation
outcomes of MPAs.

@@

As a thought experiment, imagine a region that has driven its fish
populations to near extinction and then places 100\% of its waters
inside a no-take MPA. In that setting, we would expect the
regional-scale conservation benefits to be massive. On the other hand,
imagine implementing an MPA in an area where a sedentary species has
been only lightly fished. In that setting, it is possible to create a
small net conservation loss if the concentration of fishing pressure
outside the reserve has a greater effect than the biomass buildup inside
the reserve. Within these broad bounds, numerous factors can act to
affect the regional effects of MPAs. These include the scale of adult
and larval dispersal relative to the size of the MPAs (22--25), the
strength and timing of density dependence in the population (e.g.~pre-
or post-settlement), how overfished the population would be without the
MPA, and how fishing activity and management responds to the
implementation of the MPAs (4, 8, 26--32). In addition, even for the
same total area of MPAs, the location and spacing of the MPAs can have a
profound influence on their cumulative impact through habitat and
network effects (4, 33).

@@

These simple results support the intuitive conclusion that bigger MPAs
on overfished populations produce large positive conservation outcomes.
However, when combined with other factors such as fleet dynamics, life
history, and MPA size, the conservation effects of MPAs can vary widely
in magnitude and timing, a finding corroborated by empirical evidence
(37--40) (Fig.\ref{conservation-effects}-B). As the area protected
inside MPAs increases, positive effects become more likely, but even
MPAs protecting most of a population's footprint can produce
conservation effects as low as 0\% and as high as 200\%. The degree of
depletion (i.e.~fishing pressure) has a clearer signal, with small
conservation benefits of MPAs when depletion is low, and larger effects
when depletion is high, but again even for severely depleted populations
the conservation effects of MPAs can vary widely (although nearly always
with population benefits).

@@

How can we empirically measure the regional effects of MPAs? Intuition
and bio-economic theory tells us that the regional conservation effects
of MPAs are likely to be highly context dependent across locations and
across species within locations, so careful thought must be devoted to
an empirical design to detect their effects. The perfect experiment
would involve two parallel worlds that were identical, except for the
implementation of an MPA. In world ``A'', no MPA would be implemented,
and in the facsimile world ``B'', the MPA would be implemented. Both
worlds would be tracked before and after MPA implementation, and the
conservation and fishery outcomes would be compared after treatment.
Instead of two parallel worlds, a similar experimental design would
involve random placement of MPAs.

Unfortunately, neither of these experimental designs has been
implemented, because parallel worlds do not exist and MPAs are never
placed at random. Careful causal inference requires some other means of
controlling for biases introduced by factors such as the MPA siting
process and biological spillover and concentration of fishing effort
(41, 42).

Conservation effects are typically estimated by comparing organism
densities or biomass inside MPAs to densities or biomass in selected
control sites outside MPAs, which we will refer to as response ratios.
(5) and (43) present meta-analyses of hundreds of such studies. These
results often find massively higher densities and biomasses inside MPAs
than outside (5). While response ratios are intended as a measure of the
effectiveness of reserves within their borders, these result stand in
contrast to the smaller region wide effect sizes predicted by our
simulation model. This raises the question: How reliable are density or
biomass ratios as estimators of regional conservation effects?

Control sites used in density or biomass ratios are often selected based
on abiotic or ecological traits such as habitat characteristics (41).
However, selection of control sites is further complicated by the very
spillover that MPAs are often intended to create. Export of adults or
larvae from the MPA to the ``control'' site affects their status as
controls, as does displacement of fishing effort from MPAs to control
sites. In theory, control sites far enough away to negate both
biological spillover and concentration by the fishing fleet could be
selected, but finding suitably far sites that are also appropriate
proxies for the ecological and economic context of the MPAs may be
challenging. While these concerns have been stated previously (44), the
MPA evaluation literature has by and large been unable to adequately
address them, often due to the very real challenges of acquiring the
necessary data (41).

Failure to account for these spillover effects can result in a biased
estimate of the true effect of a policy such as MPA placement (42). In
an ideal setting, the control site used in a response ratio would be a
perfect counterfactual for what would have occurred without the MPA. We
used our simulation model to approximate this scenario, calculating the
response ratio as the ratio of densities inside the MPAs relative to the
overall density from the paired simulation without MPAs multiplied by
the area of the population in an MPA (to account for the fact that we
would expect the effectiveness of a response ratio as a proxy for
regional conservation effects to scale with the size of the MPA relative
to the range of the species, Fig.\ref{density-ratio}-A). In this
idealized case, the response ratio is often a reasonable estimator for
the regional conservation effect.

What happens when the ``control'' sites are in fact affected by the
application of the MPA through biological spillover and fishing
concentration? In order to address this possibility, we calculated a new
response ratio as the mean density inside MPAs relative to the mean
density outside the MPAs (both weighted by distance from MPA borders and
scaled by MPA size). We only include simulations in which habitat and
larval dispersal rates are identical inside and outside of MPAs, to
approximate a scenario in which treatment and control sites have been
paired by ecological characteristics. Under these circumstances the
response ratio is a somewhat biased and very inaccurate estimator of the
true regional effect of MPAs.

While on average high response ratios correspond with high true regional
effects, the variance is high, so the response ratio alone cannot be
used as a reliable proxy for the true regional effect, without
accounting for the sources of biases discussed here. For higher movement
rates, response ratios were frequently near 0\% (potentially leading
stakeholders to conclude that the MPA had been ineffective), when in
fact in many cases the true effect of the MPA was highly positive. For
highly sedentary species, extremely high response ratios could create
the appearance of massive conservation gains when in fact the net effect
on the regional population has been near zero or even negative
(Fig.\ref{density-ratio}-B).

@@

\begin{figure*}%[tbhp]
  \centering
  \includegraphics[width=.9\linewidth]{figs/density-ratio.png}
  \caption{Simulated MPA-size weighted response ratios (x-axis) plotted against true regional change in biomass caused by MPAs (y-axis). Each point represents a simulation, and color represents the movement rate of fish in the simulation (expressed as a percentage of the patches a fish is likely to move in one time step). The black line shows the one:one fit line, the blue line is a linear fit of the relationship between observed response ratios and true regional change in biomass. Inside vs. Experiment simulates a true control, while Inside Vs. Outside uses a control that may be affected by the treatment.}
  \label{density-ratio}
\end{figure*}

@@

Given these challenges, what is a viable alternative for estimating the
conservation effects of MPAs? In a manner similar to (38), we use a
difference-in-difference estimator comparing densities in finfish
species targeted by fishing effort (i.e., those potentially affected by
an MPA) from those not targeted before and after MPA implementation
(42). Targeted species in the Northern Channel Islands, California,
include commercially important finfish such as California sheephead
(\emph{Semicossyphus pulcher}), and copper (\emph{Sebastes caurinus})
and blue (\emph{Sebastes mystinus}) rockfish. Each of these targeted
species was the subject of prior bio-economic modeling related to the
effects of MPAs in southern California (45, 46). But, we must note that
the analysis omits species from important invertebrate fisheries
including red urchin (\emph{Mesocentrotus franciscanus}) and spiny
lobster (\emph{Panulirus interruptus}). Non-targeted species include
garibaldi (\emph{Hypsypops rubicundus}), halfmoons (\emph{Medialuna
californiensis}), and blacksmith (\emph{Chromis punctipinnis}) (See
Table.S2 for a complete list of species). The key assumptions of this
approach are that a) within the time-frame of the model there are no
significant interaction effects between the targeted and non-targeted
species (which in fact we do not detect, see SI), and b) that in the
absence of the MPAs both the targeted and non-targeted groups of species
would have exhibited similar trends in densities. The advantage to this
approach is that given the time-frame of the model (16 years), we
believe that spillover effects of MPA placement on the non-targeted
species are likely to be much less severe than the effects of biological
spillover and fishing concentration that may bias the performance of
estimators such as a response ratios or a spatial BACI design.

We applied this difference-in-difference strategy to a large dataset in
California. We used empirical kelp forest survey data from the
Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO)
monitoring in the Northern Channel Islands with the ultimate goal of
testing the regional effects of MPAs in a real world context. A network
of MPAs covering approximately 20\% of the islands' waters was put in
place in 2003 as part of the California Marine Life Protection Act
(MLPA) (see (47), (48), (49), and (50) for information on the creation
of the MLPA). PISCO conducts visual SCUBA surveys at a large number of
rocky-reef and kelp forest sites inside and outside of MPAs throughout
the Channel Islands, producing estimates of densities of fishes that are
both targeted and non-targeted by fishing
(Fig.\ref{channel-islands}-\ref{pop-trends}).

@@

\begin{figure}%[tbhp]
  \centering
  \includegraphics[width=1\linewidth]{figs/pop-trends.pdf}
  \caption{Centered and scaled mean annual biomass densities of included targeted (red) and non-targeted (blue) fish species (faded lines) and smoothed means with 95\% confidence intervals over time (darker lines and ribbons). Vertical red dashed line shows year of MPA implementation.}
  \label{pop-trends}
\end{figure}

The raw trends in the biomass densities appear to show parallel trends
in the densities of targeted and non-targeted fishes pre-MPA, with the
targeted group diverging first positively and then negatively over time
(Fig.\ref{pop-trends}). However, these raw trends could be caused by any
number of confounding variables besides the effects of the MPAs
themselves, from changes in observer skill and sampling locations, to
shifts in environmental conditions. Using our difference-in-difference
model in an effort to control for these factors, we found no significant
difference in the biomass densities of targeted and non-targeted fishes
pre-MPA (before 2003). Following the implementation of the MPAs in 2003,
we see evidence of an increasing trend in densities of targeted relative
to non-targeted fish, reaching a peak value in 2014. After 2014,
densities of targeted species appear to have declined relative to what
we would expect based on the non-targeted group (Fig.\ref{did-plot}).

\begin{figure}%[tbhp]
  \centering
  \includegraphics[width=1\linewidth]{figs/did_plot.png}
  \caption{Estimated divergence in biomass densities of targeted and non-targeted fishes throughout the Channel Islands (integrated across inside and outside of MPAs). MPAs are implemented in 2003 (red dashed line). Colored cells in the background indicate the frequency of MPA conservation effects predicted by our simulation modeling for MPAs similar in design and species composition to the Channel Islands in each year.}
  \label{did-plot}
\end{figure}

@@

We find regional conservation gains from MPAs in 94\% of our simulations
(Fig.\ref{conservation-effects}). MPAs covering less than 25\% of the
population range were unlikely to produce regional conservation gains
above 10\% unless the stock would have been severely overfished without
the MPA. Given that few marine species on the planet have more than 25\%
of their entire range protected in MPAs, this suggests that large
regional gains are unlikely to be common unless the overall extent of
MPAs grows globally. Large MPAs protecting highly depleted stocks almost
always produced large regional conservation gains, as should be
expected.

The median conservation effect of 36\% stands in contrast to the large
within-MPA effects on biomass densities reported by (51) and (5). This
difference is due in part to the fact that within-MPA effects are likely
to be much larger than regional-MPA effects, but may also reflect the
many challenges of translating response ratios into regional
conservation effects. Our simulation analysis shows that without proper
controls (in particular for the effects of fleet redistribution),
response ratios well over 100\% are entirely plausible even when the
actual conservation effect is much lower (and \emph{vice versa}).

@@

Our simulation analysis suggests that this result should perhaps be
expected. The Channel Islands MPAs cover approximately 20\% of the
waters in the Channel Islands, and while formal stock assessments are
not available for many of the targeted species in our analysis, what
evidence we have suggests that, as a group, these fish are on average
not heavily overfished. Some species, such as California sheephead and
blue rockfish were likely below target levels during the period (52,
53), but projections based upon the overall average response across all
species will likely suggest modest benefits even if a subset of species
could experience much larger population gains. Harkening back to our
simulations, we expect the average percentage difference in densities of
targeted species with and without MPAs to be modest
(Fig.\ref{did-plot}). Effects of this size are likely to be challenging
to detect empirically given the large natural variation of marine
ecosystems (especially temperate reefs) and the observation error
inherent in infrequent annual survey programs such as those provided by
PISCO.

@@

While the MPA effect estimated by our model was not significant in any
one year, the positive trend from 2005 to 2014 is evident, as is the
sharp decline from 2015-2017. To what should we attribute this apparent
shift? We cannot reject the parallel trends assumption between the
targeted and non-targeted species in the years before the MPAs
(Fig.\ref{did-plot}). But, the Channel Islands region (and the entire
West coast of the USA) experienced a dramatic `marine heatwave'
beginning in 2014 and persisting through 2016, resulting in part in
extremely elevated water temperatures throughout the region (54).
Biogeographic differences in the distributions of targeted and
non-targeted species may confound the observed effects of MPAs. Many of
the non-targeted species have warm thermal affinities and have increased
in numbers since the heatwave (55). The targeted group is made up mostly
of fishes with cold-water affinities, such as members of the genus
\emph{Sebastes}. As such, we hypothesize that the recent evidence for a
decline in densities of targeted species is due to environmental
conditions that disproportionately affect the targeted group (and not
for example due to concentrated fishing pressure outside the reserves).
This hypothesis is supported by the sharp declines in densities of
targeted species seen inside the MPAs themselves (see Fig.S27-S29),
suggesting a driver other than fishing may be at play.

@@

Once simulation models have been used to help design an MPA, how can
users evaluate whether it is achieving their objectives? Response ratios
are commonly used as evidence for conservation outcomes of MPAs (5), but
as suggested in (41) and (42) and demonstrated here, without careful
attention to the design of control sites (accounting for example for the
displacement of effort by MPAs), response ratios may be highly
unreliable estimators of regional MPA effects. As (42) suggests, there
are many potential alternatives for estimating the effects of MPAs that
better account for the challenges of causal inference. We applied one
such approach here (a difference-in-difference estimator), and yet were
still unable to reach robust conclusions as to the effect of MPAs on the
density of targeted species in the Channel Islands, due to the likely
small size of the true effect relative to the strength and variability
of environmental drivers.
